### 2020-02-10 thesis/content/chapter-introduction.tex
continued related works chapter

The field of \ac{ml} on graph-structured data has applications in many domains due to the general expressive power of graphs.
Three common types of graph~\ac{ml} problems are
\begin{enumerate}[label=\textbf{\arabic*.}]
	\item \textbf{Link prediction:}
		A graph with an incomplete edge set is given and the missing edges have to be predicted.
		The generation of friendship suggestions in a social network is a typical example for this.
	\item \textbf{Vertex classification \& regression:}
		Here a class or a score has to be predicted for each vertex of a graph.
		In social graphs this corresponds to the prediction of properties of individuals.
		Another example is the prediction of the amount of traffic at the intersections of a street network.
	\item \textbf{Graph classification \& regression:}
		In this final problem type a single global class or continuous value has to be predicted for an input graph.
		The canonical example for this is the prediction of properties of molecule graphs, e.g.\ the toxicity or solubility of a chemical.
\end{enumerate}
In this thesis we will focus on the last problem type, \ac{gcr}.
A \ac{ml} method for this problem has to accept graphs of varying size and should be permutation invariant wrt.\ the vertices.
Those requirements are not met by the commonly used learners that only accept fixed-size feature vectors as their input, e.g.\ \acp*{lm}, \acp{svm} or \acp{mlp}.

>>>>

The field of \ac{ml} on graph-structured data has applications in many domains due to the general expressive power of graphs.
Three common types of graph~\ac{ml} problems are
\begin{enumerate}[label=\textbf{\arabic*.}]
	\item \textbf{Link prediction:}
		A graph with an incomplete edge set is given and the missing edges have to be predicted.
		The generation of friendship suggestions in a social network is a typical example for this.
	\item \textbf{Vertex classification \& regression:}
		Here a class or a score has to be predicted for each vertex of a graph.
		In social graphs this corresponds to the prediction of properties of individuals.
		Another example is the prediction of the amount of traffic at the intersections of a street network.
	\item \textbf{Graph classification \& regression:}
		In this final problem type a single global class or continuous value has to be predicted for an input graph.
		The canonical example for this is the prediction of properties of molecule graphs, e.g.\ the toxicity or solubility of a chemical.
\end{enumerate}
In this thesis we will focus on the last problem type, \acf{gcr}.
An \ac{ml} method for this problem has to accept graphs of varying size and should be permutation invariant wrt.\ the vertices.
Those requirements are not met by the commonly used learners that only accept fixed-size feature vectors as their input, e.g.\ \acp*{lm}, \acp{svm} or \acp{mlp}.

### 2020-02-10 thesis/content/chapter-related-work.tex
implemented evaluation logging and continuation
- the evaluation loop now logs all observed train, validation and test metrics
- it also stores the required metadata to continue an interrupted evaluation
continued related works chapter

Before combining \ac{lta} and \ac{gcr} as described in \cref{sec:intro:goals}, we first give an overview of the state-of-the-art in both fields of research.
This is done in three steps:
\begin{enumerate}
	\item We begin with an overview of the existing \ac{lta} methods for set inputs.
	\item Before the existing \ac{gcr} methods are described, we look at the graph theoretical fundamentals on which those methods are based.
	\item Using the previously introduced fundamentals, the current approaches to tackle the \ac{gcr} problem are then described.
\end{enumerate}

@@

\section{Graph Theoretical Fundamentals}%
\label{sec:related:theory}

\subsection{Graph Isomorphism Checks}%
\label{sec:related:theory:wl}

\subsection{Spectral Graph Theory}%
\label{sec:related:theory:spectral}

\section{Graph Regression and Classification}%
\label{sec:related:gcr}

The existing approaches to tackle the \ac{gcr} problem can be categorized into three main families:
\begin{enumerate*}
	\item Explicit graph embeddings,
	\item graph kernels and
	\item graph neural networks
\end{enumerate*}.
We will now look at the characteristics of those families and give an overview of specific methods.

@@

\subsubsection{Fingerprint embeddings}

@@

\subsubsection{Skip-gram inspired embeddings}

@@

\paragraph{Vertex embeddings}
Skip-gram embeddings can be  na{\"\i}vely extended to graphs by realizing that \texttt{word2vec} effectively already is a vertex embedding method for linear graphs in which $C_k(v)$ is simply the $k$-neighborhood of the vertex/word $v$.
The problem with this na{\"\i}ve extension is that the sizes of $k$-neighborhoods in arbitrary graphs can be much larger and often tend to grow exponentially with $k$.
To deal with this computational problem the so-called DeepWalk~\cite{Perozzi2014} and \texttt{node2vec}~\cite{Grover2016} methods perform random walks of fixed length to effectively take samples from the neighborhood of vertices.
Both methods only differ in the transition matrix that is used for the random walk.
Another difference of DeepWalk and \texttt{node2vec} compared to \texttt{word2vec} is the so-called \textit{feature space symmetry} which states that the context root interpretation ($f$) of a vertex should be symmetric to its context element interpretation ($f_C$), i.e.\ $f = f_C$.
The combination of random walk context sampling and and the feature space symmetry assumption can be used to compute vertex embeddings even for very large graphs.

\paragraph{Graph embeddings}
Skip-gram methods can not only be used for vertex embeddings but also to embed entire graphs.
One way to do this is via the \texttt{graph2vec}~\cite{Narayanan2017} method.
It is inspired by \texttt{doc2vec}~\cite{Le2014} which in turn is based on \texttt{word2vec}.
\texttt{graph2vec} gets a set of graphs $\mathbb{G} = \{ G_1, \dots G_N \}$ as input and outputs graph embeddings $f(G_1), \dots, f(G_N)$.
While in \texttt{word2vec} every word can be a context root as well as a context element, \texttt{graph2vec} uses the graphs $\mathbb{G}$ as context roots.
The context $C_k(G_i)$ of a graph $G_i = (\mathcal{V}_i, \mathcal{E}_i)$ is defined as
\begin{equation}
	C_k(G_i) := \bigcup_{\mathclap{v_j \in \mathcal{V}_i}}\, {\left\{ {\textrm{WL}_l(G_i)}_j \right\}}_{l = 0}^k % chktex 21
	\text{ with } \textrm{WL}_l(G) := \text{\acs{wl} colors after $l$ iter.\ on $G$.}
\end{equation}
Intuitively this context can be understood as the set of \acs{wl}-distinguishable subgraphs of $G_i$ with diameter $\leq 2k$.
Since \ac{wl} is used to identify distinct subgraphs, \texttt{graph2vec} can only be applied to graphs with discrete vertex labels.
Using the previous definitions, the context root embedding function has the signature $f: \mathbb{G} \to \mathbb{R}^d$ while the context element embedding function is of type $f_C: \bigcup_{G_i \in \mathbb{G}} C_k(G_i) \to \mathbb{R}^d$.
To find those embeddings the \texttt{word2vec} objective from \cref{eq:related:gcr:embed:word2vec} is reused.
Analogous to \texttt{word2vec}, \texttt{graph2vec} therefore embeds graphs that share subgraphs close to each other, whereas graphs that do not share substructures tend to be embedded further away from each other.

@@

Instead of explicity mapping a graph or vertices into a vector space, one can also do so implicity by employing the kernel trick.
There is a large variety of so-called \acp{gk} to do this.
\acp{gk} can be used in combination with any kernel method to solve the \ac{gcr} problem.

>>>>

Before combining \ac{lta} and \ac{gcr} as described in \cref{sec:intro:goals}, we first give an overview of the state-of-the-art in both fields of research.
This is done in three steps:
\begin{enumerate}
	\item We begin with an overview of the existing \ac{lta} methods for unstructured inputs.
	\item Then we look at the domain of structured inputs.
		To solve the \ac{gcr} problem, relevant graphs characteristics have to be defined in order to determine the similarity and dissimilarity of graphs.
		We will look at two common approaches for graph characterization: The \acl{wl} algorithm and the notion of graph spectra.
	\item Using the described graph characterization approaches, a brief overview of current \ac{gcr} methods will then be given.
\end{enumerate}

@@

\section{Graph Characterization}%
\label{sec:related:character}

To classify or score a graph, it first needs to be characterized by a set of relevant properties.
Two commonly used graph characterization approaches are the so called \acf{wl} colors and the notion of the graph spectrum.
Both will now be briefly introduced since they form the foundation of most current \ac{gcr} methods.

\subsection{Weisfeiler-Lehman Graph Colorings}%
\label{sec:related:character:wl}

The Weisfeiler-Lehman algorithm characterizes a graph $G = (\mathcal{V}, \mathcal{E})$ by assigning discrete labels $c \in \mathcal{C}$, called \textit{colors}, to vertex $k$-tuples $(v_1, \dots, v_k) \in \mathcal{V}^k$, where $k \in \mathbb{N}$ is the freely choosable \textit{\ac{wl}-dimension}.
A mapping $\chi: \mathcal{V}^k \to \mathcal{C}$ is called a coloring of $G = (\mathcal{V}, \mathcal{E})$.

\paragraph{The 1-dimensional \acs{wl} algorithm}
For the most simple case of $k = 1$, a color is assigned to each vertex of a graph.

\subsection{Spectral Graph Theory}%
\label{sec:related:character:spectral}

\section{Graph Classification and Regression}%
\label{sec:related:gcr}

The existing approaches to tackle the \ac{gcr} problem can be categorized into three main families:
\begin{enumerate*}
	\item Explicit graph embeddings,
	\item graph kernels and
	\item graph neural networks
\end{enumerate*}.
We will now look at the characteristics of those families and give a brief overview of specific methods.

@@

\subsubsection{Fingerprint Embeddings}

@@

\subsubsection{Skip-gram inspired Embeddings}

@@

\paragraph{Vertex Embeddings}
Skip-gram embeddings can be  na{\"\i}vely extended to graphs by realizing that \texttt{word2vec} effectively already is a vertex embedding method for linear graphs in which $C_k(v)$ is simply the $k$-neighborhood of the vertex/word $v$.
The problem with this na{\"\i}ve extension is that the sizes of $k$-neighborhoods in arbitrary graphs can be much larger and often tend to grow exponentially with $k$.
To deal with this computational problem the so-called DeepWalk~\cite{Perozzi2014} and \texttt{node2vec}~\cite{Grover2016} methods perform random walks of fixed length to effectively take samples from the neighborhood of vertices.
Both methods only differ in the transition matrix that is used for the random walk.
Another difference of DeepWalk and \texttt{node2vec} compared to \texttt{word2vec} is the so-called \textit{feature space symmetry} which states that the context root interpretation ($f$) of a vertex should be symmetric to its context element interpretation ($f_C$), i.e.\ $f = f_C$.
The combination of random walk context sampling and and the feature space symmetry assumption can be used to compute vertex embeddings even for very large graphs.

\paragraph{Graph Embeddings}
Skip-gram methods can not only be used for vertex embeddings but also to embed entire graphs.
One way to do this is via the \texttt{graph2vec}~\cite{Narayanan2017} method.
It is inspired by \texttt{doc2vec}~\cite{Le2014} which in turn is based on \texttt{word2vec}.
\texttt{graph2vec} gets a set of graphs $\mathbb{G} = \{ G_1, \dots G_N \}$ as input and outputs graph embeddings $f(G_1), \dots, f(G_N)$.
While in \texttt{word2vec} every word can be a context root as well as a context element, \texttt{graph2vec} uses the graphs $\mathbb{G}$ as context roots.
The context $C_k(G_i)$ of a graph $G_i = (\mathcal{V}_i, \mathcal{E}_i)$ is defined as
\begin{equation}
	C_k(G_i) := \bigcup_{\mathclap{v_j \in \mathcal{V}_i}}\, {\left\{ {\textrm{WL}_l(G_i)}_j \right\}}_{l = 0}^k % chktex 21
	\text{ with } \textrm{WL}_l(G) := \text{$l$ iter.\ \acs{wl}-1 coloring of $G$.}
\end{equation}
Intuitively this context can be understood as the set of \acs{wl}-distinguishable subgraphs of $G_i$ with diameter $\leq 2k$.
Since \ac{wl} is used to identify distinct subgraphs, \texttt{graph2vec} can only be applied to graphs with discrete vertex labels.
Using the previous definitions, the context root embedding function has the signature $f: \mathbb{G} \to \mathbb{R}^d$ while the context element embedding function is of type $f_C: \bigcup_{G_i \in \mathbb{G}} C_k(G_i) \to \mathbb{R}^d$.
To find those embeddings the \texttt{word2vec} objective from \cref{eq:related:gcr:embed:word2vec} is reused.
Analogous to \texttt{word2vec}, \texttt{graph2vec} therefore embeds graphs that share subgraphs close to each other, whereas graphs that do not share substructures tend to be embedded further away from each other.

@@

Instead of explicitly mapping a graph or vertices into a vector space, one can also do so implicitly by employing the kernel trick.
There is a large variety of so-called \acp{gk} to do this.
\acp{gk} can be used in combination with any kernel method, typically a \ac{svm}, to solve the \ac{gcr} problem.
While there is a large variety of different \acp{gk}, we will focus on those that are based on the previously described family of \ac{wl} algorithms.

\paragraph{\ac{wl} subtree kernel}

\paragraph{\ac{wl} shortest path kernel}

\paragraph{Higher dimensional \ac{wl} kernels}
